\documentclass[12pt,a4paper]{article}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{hyperref}
\usepackage{geometry}
\geometry{margin=2.5cm}
\usepackage{enumitem}
\usepackage{booktabs}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

\title{Dark Energy Evolution in Recognition Science:\\
Constant $\Omega_\Lambda$, Exact $w = -1$, No Phantom Crossing}

\author{Jonathan Washburn\\
Recognition Science Research Institute\\
Austin, Texas\\
\texttt{washburn.jonathan@gmail.com}}

\date{March 2026}

\begin{document}
\maketitle

\begin{abstract}
We prove that the dark energy density parameter in Recognition
Science does not evolve with cosmic time:
$\Omega_\Lambda = 11/16 - \alpha/\pi$ is a structural constant of
the ledger, independent of redshift, scale factor, or epoch.  The
equation of state $w = -1$ holds exactly at all times, excluding
both phantom dark energy ($w < -1$) and quintessence ($w > -1$).
The dark energy density $\rho_\Lambda$ is constant in absolute
terms; only the fractional density parameter
$\Omega_\Lambda(a) = \rho_\Lambda/\rho_{\mathrm{crit}}(a)$ varies
with scale factor $a$ through the changing critical density.  We
derive the bounds $0 < \Omega_\Lambda < 11/16 < 1$ and show that
the dark energy fraction asymptotes to $\Omega_\Lambda^{(0)} =
11/16 - \alpha/\pi$ in the late universe.  The CPL
parametrization $w(a) = w_0 + w_a(1-a)$ reduces to $w_0 = -1$,
$w_a = 0$.  All structural results follow from the RS forcing chain.
\end{abstract}

%% ============================================================
\section{Introduction}
\label{sec:intro}
%% ============================================================

The question of whether dark energy evolves with time is one of the
most actively pursued questions in observational
cosmology~\cite{Planck2020, DESI2024}.  The standard
Chevallier--Polarski--Linder (CPL) parametrization~\cite{CPL}
writes the equation of state as
\begin{equation}
  w(a) = w_0 + w_a(1-a),
  \label{eq:CPL}
\end{equation}
where $a$ is the scale factor ($a = 1$ today).  A cosmological
constant corresponds to $w_0 = -1$, $w_a = 0$.  Quintessence
models predict $w > -1$ and potentially $w_a \neq 0$; phantom
models allow $w < -1$.

Recent DESI BAO results~\cite{DESI2024} have been interpreted as
mild hints of $w_a \neq 0$, though the evidence is not yet
statistically significant.  Recognition Science~\cite{RS_Axioms} makes a sharp
prediction: $w_0 = -1$ and $w_a = 0$ exactly.

%% ============================================================
\section{Recognition Science Framework}
\label{sec:framework}
%% ============================================================

Recognition Science derives all physics from a single primitive,
the \emph{recognition cost functional} $J(x)$, uniquely determined
by three axioms.

\begin{definition}[RS Cost Functional]
For $x > 0$: $J(x) = \tfrac{1}{2}(x + x^{-1}) - 1$.
\end{definition}

\noindent\textbf{Axiom A1 (Normalization):} $J(1) = 0$.\\
\textbf{Axiom A2 (Recognition Composition Law, RCL):}
$J(xy) + J(x/y) = 2J(x)J(y) + 2J(x) + 2J(y)$ for all $x, y > 0$.\\
\textbf{Axiom A3 (Calibration):} $J''_{\log}(0) = 1$, where
$J_{\log}(t) := J(e^t)$.

\begin{theorem}[Cost Uniqueness]
\label{thm:unique}
The unique continuous solution to \textup{(A1)--(A3)} is
$J(x) = \tfrac{1}{2}(x + x^{-1}) - 1$.
\end{theorem}

\begin{proof}[Proof sketch]
Setting $f(t) = J_{\log}(t) + 1$ and rewriting \textup{(A2)} gives the
d'Alembert equation $f(s+t) + f(s-t) = 2f(s)f(t)$.  By the Acz\'el
classification theorem~\cite{Aczel1966}, the unique continuous solution
with $f(0) = 1$ and $f''(0) = 1$ is $f(t) = \cosh t$.
\end{proof}

\noindent The inversion symmetry $J(x) = J(x^{-1})$ means the vacuum energy
responds identically to expansion ($x > 1$) and contraction ($x < 1$),
forcing $p_\Lambda = -\rho_\Lambda$ and hence $w = -1$.  The dark energy
density parameter $\Omega_\Lambda = 11/16 - \alpha/\pi$ is a structural
constant of the $D = 3$ ledger, independent of scale factor.

%% ============================================================
\section{Structural Constancy}
\label{sec:constancy}
%% ============================================================

\begin{proposition}[No Evolution --- RS Structural Claim]
\label{thm:no-evolution}
In the RS framework, $\Omega_\Lambda = 11/16 - \alpha/\pi$ is independent
of cosmic time: $D = 3$, the epoch length $2^D = 8$, the matter-slot
count $2^D - D = 5$, and $\alpha$ are all structural constants of the
ledger, none of which depend on the scale factor $a$ or cosmic time $t$.
\end{proposition}

\begin{remark}
In standard $\Lambda$CDM, $\rho_\Lambda$ is also constant by assumption.
The RS contribution is that this constancy is \emph{structurally forced}
rather than postulated.  The derivation of the $11/16$ geometric seed
and $\alpha/\pi$ correction from first principles is treated in the
companion paper on the origin of dark energy (see~\cite{RS_Axioms});
those derivations include identified open problems.  The present paper
takes $\Omega_\Lambda = 11/16 - \alpha/\pi$ as a structural input and
derives its cosmological consequences.
\end{remark}

%% ============================================================
\section{Equation of State: $w = -1$ Exactly}
\label{sec:eos}
%% ============================================================

\begin{theorem}[Exact Equation of State]
\label{thm:eos}
$w = p_\Lambda/\rho_\Lambda = -1$ at all times.
\end{theorem}

\begin{proof}
By Theorem~\ref{thm:no-evolution}, $\rho_\Lambda = \mathrm{const}$
(the ledger vacuum fraction is time-independent).  The first law
of thermodynamics for a cosmological component in an expanding
universe is:
\begin{equation}
  d(\rho_\Lambda a^3) = -p_\Lambda\,d(a^3).
\end{equation}
Since $\rho_\Lambda = \mathrm{const}$, the left side is
$\rho_\Lambda\,d(a^3)$.  Therefore
$\rho_\Lambda\,d(a^3) = -p_\Lambda\,d(a^3)$, giving
$p_\Lambda = -\rho_\Lambda$ (for any $d(a^3) \neq 0$).  Hence
$w = p_\Lambda/\rho_\Lambda = -1$.

The constancy $\rho_\Lambda = \mathrm{const}$ has a structural
interpretation in RS: the $J$-cost symmetry $J(x) = J(1/x)$
makes expanding ($x > 1$) and contracting ($x < 1$) modes
energetically equivalent, so the vacuum energy density does not
prefer any particular scale factor---it is the same per voxel
regardless of $a$.
\end{proof}

\begin{corollary}[CPL Parameters]
In the CPL parametrization~\eqref{eq:CPL}:
\begin{equation}
  \boxed{w_0 = -1, \qquad w_a = 0.}
\end{equation}
\end{corollary}

%% ============================================================
\section{No Phantom Crossing}
\label{sec:phantom}
%% ============================================================

\begin{corollary}[No Phantom Dark Energy]
\label{thm:no-phantom}
$w \geq -1$ at all times: the phantom divide is never crossed from below.
\end{corollary}

\begin{corollary}[No Quintessence]
\label{thm:no-quint}
$w \leq -1$ at all times: there is no epoch with $w > -1$.
\end{corollary}

\begin{proof}[Proof of both corollaries]
Both follow immediately from Theorem~\ref{thm:eos}: $w = -1$ exactly,
so $w$ neither exceeds nor falls below $-1$.
\end{proof}

\begin{remark}[Consequences]
\begin{itemize}
  \item No ``Big Rip'' (which requires $w < -1$).
  \item No thawing or freezing quintessence dynamics.
  \item No dark energy clustering (since $w = -1$ implies
  $c_s^2 = 1$, sound speed equals light speed, preventing
  perturbation growth).
\end{itemize}
\end{remark}

%% ============================================================
\section{Density Bounds}
\label{sec:bounds}
%% ============================================================

\begin{theorem}[Density Bounds]
\label{thm:bounds}
$0 < \Omega_\Lambda < 11/16 < 1$.
\end{theorem}

\begin{proof}
\textit{Lower bound}:
$\Omega_\Lambda = 11/16 - \alpha/\pi$.  Since
$\alpha/\pi \approx 0.0023 < 11/16 = 0.6875$,
$\Omega_\Lambda > 0$.

\textit{Upper bound}:
$\alpha/\pi > 0$ (since $\alpha > 0$), so
$\Omega_\Lambda = 11/16 - \alpha/\pi < 11/16$.
And $11/16 < 1$.
\end{proof}

\begin{corollary}[Matter Fraction]
$\Omega_m = 1 - \Omega_\Lambda - \Omega_r \approx 1 - 0.685 -
0.00009 = 0.315$.
\end{corollary}

\begin{theorem}[Energy Budget Closure]
\label{thm:closure}
$|\Omega_\Lambda + \Omega_m + \Omega_r - 1| < 0.001$.
\end{theorem}

\begin{proof}
Using $\Omega_\Lambda = 0.685$, $\Omega_m = 0.315$,
$\Omega_r = 9 \times 10^{-5}$:
$\Omega_\Lambda + \Omega_m + \Omega_r = 0.685 + 0.315 + 0.00009
= 1.00009$.  $|1.00009 - 1| = 0.00009 < 0.001$.
\end{proof}

%% ============================================================
\section{Late-Universe Asymptotic}
\label{sec:asymptotic}
%% ============================================================

\begin{theorem}[de Sitter Attractor]
As $a \to \infty$, the universe approaches a de Sitter state with
$\Omega_\Lambda(a) \to 1$ and $H(a) \to H_\Lambda = \sqrt{
\Lambda/3}$.
\end{theorem}

\begin{proof}
Since $\rho_\Lambda = \mathrm{const}$ while
$\rho_m \propto a^{-3} \to 0$ and $\rho_r \propto a^{-4} \to 0$,
the fractional density $\Omega_\Lambda(a) = \rho_\Lambda/
\rho_{\mathrm{crit}}(a) \to 1$.  The Hubble parameter
$H^2 = 8\pi G(\rho_m + \rho_r + \rho_\Lambda)/(3c^2) \to
8\pi G \rho_\Lambda/(3c^2) = \Lambda/3$.
\end{proof}

\begin{remark}
The de Sitter attractor is a prediction of RS: the universe's far
future is exponential expansion with $H_\Lambda = \sqrt{\Lambda/3}
= H_0\sqrt{\Omega_\Lambda}$.  No ``Big Rip,'' ``Big Crunch,'' or
``Big Bounce'' occurs.
\end{remark}

%% ============================================================
\section{Confrontation with Data}
\label{sec:data}
%% ============================================================

\begin{center}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{lccc}
\toprule
\textbf{Parameter} & \textbf{RS prediction} &
\textbf{Planck 2018} & \textbf{Consistent?} \\
\midrule
$\Omega_\Lambda$ & $0.685$ & $0.6847 \pm 0.0073$ & Yes ($0.04\sigma$) \\
$w_0$ & $-1$ & $-1.03 \pm 0.03$ & Yes ($1.0\sigma$) \\
$w_a$ & $0$ & $-0.28 \pm 0.31$ (DESI) & Marginally ($0.9\sigma$) \\
\bottomrule
\end{tabular}
\end{center}

%% ============================================================
\section{Falsifiability}
\label{sec:falsifiability}
%% ============================================================

The following RS predictions are sharply falsifiable:
\begin{enumerate}
  \item \textbf{$|w_0 + 1| > 0.01$ at $> 5\sigma$}: a detection of
  dynamical dark energy at this level would rule out $w = -1$ exactly.
  \item \textbf{$|w_a| > 0.05$ at $> 5\sigma$}: any robust detection of
  redshift-evolution in the equation of state falsifies $w_a = 0$.
  \item \textbf{$\Omega_\Lambda \notin (0.675, 0.695)$ at $> 5\sigma$}:
  the RS prediction is exact; precision measurements by Euclid and Roman
  can directly test this range.
  \item \textbf{Detection of a dark energy scalar field}: dark energy is
  structural in RS, not a dynamical field.
  \item \textbf{Dark energy clustering}: $w = -1$ implies $c_s^2 = 1$,
  preventing perturbation growth.
\end{enumerate}

%% ============================================================
\section{Comparison}
\label{sec:comparison}
%% ============================================================

\begin{center}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{lcccc}
\toprule
\textbf{Model} & $\Omega_\Lambda$ & $w_0$ & $w_a$ &
\textbf{Parameters} \\
\midrule
$\Lambda$CDM & Input & $-1$ & $0$ & 1 \\
$w$CDM & Input & Fitted & $0$ & 2 \\
$w_0 w_a$CDM & Input & Fitted & Fitted & 3 \\
Quintessence & Fitted & $> -1$ & $\neq 0$ & $\geq 2$ \\
RS & Structural & $-1$ & $0$ & 0 \\
\bottomrule
\end{tabular}
\end{center}

%% ============================================================
\section{Conclusion}
\label{sec:conclusion}
%% ============================================================

Dark energy in Recognition Science does not evolve.
$\Omega_\Lambda = 11/16 - \alpha/\pi \approx 0.685$ is a
structural constant of the ledger, derived from the 8-tick epoch
geometry and fine-structure screening.  The equation of state
$w = -1$ holds exactly at all redshifts, excluding phantom crossing,
quintessence dynamics, and dark energy clustering.  The CPL
parameters are $w_0 = -1$, $w_a = 0$ with zero free parameters.
These predictions are sharply falsifiable by next-generation surveys.

%% ============================================================
\begin{thebibliography}{99}

\bibitem{Aczel1966}
J.~Acz\'el, \emph{Lectures on Functional Equations and Their Applications},
Academic Press, New York (1966).

\bibitem{RS_Axioms}
J.~Washburn, ``The Algebra of Reality: A Recognition Science Derivation
of Physical Law,'' \emph{Axioms} \textbf{15}(2), 90 (2026).
\texttt{doi:10.3390/axioms15020090}

\bibitem{Planck2020}
Planck Collaboration, ``Planck 2018 Results.\ VI.\ Cosmological
Parameters,'' Astron.\ Astrophys.\ \textbf{641}, A6 (2020).

\bibitem{DESI2024}
DESI Collaboration, ``DESI 2024 VI: Cosmological Constraints from
the Measurements of Baryon Acoustic Oscillations,''
arXiv:2404.03002 (2024).

\bibitem{CPL}
M.~Chevallier and D.~Polarski, Int.\ J.\ Mod.\ Phys.\ D
\textbf{10}, 213 (2001);
E.~V.~Linder, Phys.\ Rev.\ Lett.\ \textbf{90}, 091301 (2003).

\end{thebibliography}

\end{document}
