\documentclass[12pt,a4paper]{article}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{hyperref}
\usepackage{geometry}
\geometry{margin=2.5cm}
\usepackage{enumitem}
\usepackage{booktabs}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

\title{Proton Stability from Recognition Science:\\
Color-Singlet Protection, No Grand Unification,\\
and Exact Baryon-Number Conservation}

\author{Jonathan Washburn\\
Recognition Science Research Institute\\
Austin, Texas\\
\texttt{washburn.jonathan@gmail.com}}

\date{March 2026}

\begin{document}
\maketitle

\begin{abstract}
We derive absolute proton stability from the Recognition
Science (RS) framework.  Three independent structural
mechanisms protect the proton: (1)~color confinement from
the ledger's face-pair structure forbids decay to free
quarks; (2)~the gauge group
$SU(3) \times SU(2) \times U(1)$ is derived from
$D = 3$ cube geometry and does not unify into a larger
group at any scale, so no $X$ or $Y$ bosons exist to
mediate baryon-number violation; (3)~baryon number is an
exact ledger quantum number, with $B$-violating
configurations incurring divergent $J$-cost.  The
prediction is $\tau_p = \infty$: the proton is absolutely
stable.  The current Super-Kamiokande bound
$\tau(p \to e^+\pi^0) > 2.4 \times 10^{34}$~years is
automatically satisfied.  Any observed proton decay event
would falsify RS.
\end{abstract}

%% ============================================================
\section{Recognition Science Framework}
\label{sec:framework}
%% ============================================================

Recognition Science derives all physics from a single primitive,
the \emph{recognition cost functional} $J(x)$, uniquely determined
by three axioms.

\begin{definition}[RS Cost Functional]
For $x > 0$: $J(x) = \tfrac{1}{2}(x + x^{-1}) - 1$.
\end{definition}

\noindent\textbf{Axiom A1 (Normalization):} $J(1) = 0$.\\
\textbf{Axiom A2 (Recognition Composition Law, RCL):}
$J(xy) + J(x/y) = 2J(x)J(y) + 2J(x) + 2J(y)$ for all $x, y > 0$.\\
\textbf{Axiom A3 (Calibration):} $J''_{\log}(0) = 1$, where
$J_{\log}(t) := J(e^t)$.

\begin{theorem}[Cost Uniqueness]
\label{thm:unique}
The continuous solution to \textup{(A1)--(A3)} is unique:
$J(x) = \tfrac{1}{2}(x + x^{-1}) - 1$.
\end{theorem}

\begin{proof}[Proof sketch]
Setting $f(t) = J_{\log}(t) + 1$ and rewriting \textup{(A2)} gives the
d'Alembert equation $f(s+t) + f(s-t) = 2f(s)f(t)$.  By the Acz\'el
classification theorem~\cite{Aczel1966}, the unique continuous solution
with $f(0) = 1$ and $f''(0) = 1$ is $f(t) = \cosh t$.
\end{proof}

\noindent Three independent structural features of RS guarantee proton
stability: (1)~color confinement (ledger color singlet requirement);
(2)~absence of gauge unification (gauge group $SU(3)\times SU(2)\times U(1)$
derived from $D=3$ cube geometry, not descended from a GUT); and
(3)~exact baryon-number conservation (double-entry ledger).

%% ============================================================
\section{Introduction}
\label{sec:intro}
%% ============================================================

Proton stability is one of the sharpest discriminators
between the Standard Model (SM) and its extensions in the
context of Recognition Science~\cite{RS_Axioms}.  In the
SM, baryon number $B$ is an accidental global symmetry---it
is conserved classically but violated by non-perturbative
electroweak processes (sphalerons)~\cite{tHooft1976}.  Grand
Unified Theories (GUTs) predict proton decay as a signature
of quark--lepton unification at high
energies~\cite{GeorgiGlashow1974}.

The experimental status:
\begin{center}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{@{}llc@{}}
\toprule
\textbf{Channel} & \textbf{Experiment} &
\textbf{$\tau$ bound (yr, 90\% CL)} \\
\midrule
$p \to e^+\pi^0$ & Super-K~\cite{SK2024}
& $> 2.4 \times 10^{34}$ \\
$p \to \bar{\nu} K^+$ & Super-K
& $> 5.9 \times 10^{33}$ \\
$p \to \mu^+\pi^0$ & Super-K
& $> 1.6 \times 10^{34}$ \\
$p \to e^+\eta$ & Super-K
& $> 4.2 \times 10^{33}$ \\
\bottomrule
\end{tabular}
\end{center}

These null results have already excluded minimal $SU(5)$
and constrain SUSY GUTs.  RS predicts that all proton
decay searches will continue to yield null results
indefinitely: $\tau_p = \infty$.

%% ============================================================
\section{Mechanism 1: Color Confinement}
\label{sec:confinement}
%% ============================================================

\begin{definition}[Color Charge from Cube Geometry]
\label{def:color}
In RS, the $D = 3$ cube has 3 face-pairs
($N_c = F/2 = 2D/2 = D = 3$).  Each face-pair
defines a color degree of freedom, giving $SU(3)$ color.
Color confinement is the requirement that all physical
states are color singlets---balanced entries in the
recognition ledger.
\end{definition}

\begin{lemma}[No Free Quarks]
\label{lem:no-quarks}
Free quarks carry unbalanced color charge, which
corresponds to an unbalanced ledger entry with divergent
$J$-cost.  Physical states must be color singlets.
\end{lemma}

\begin{proposition}[Proton as Minimal Color-Singlet Baryon --- Empirical Fact]
\label{thm:minimal}
The proton is the lightest color-singlet state with
baryon number $B = 1$.
\end{proposition}

\begin{proof}
The quark content of the proton is $uud$, forming a color
singlet $\epsilon_{ijk} u^i u^j d^k$ with $B = 1$.  The
neutron ($udd$, $m_n = 939.565~\mathrm{MeV}$) is heavier
than the proton ($m_p = 938.272~\mathrm{MeV}$) by
$1.293~\mathrm{MeV}$.  All other baryons ($\Lambda$,
$\Sigma$, $\Xi$, $\Omega$, etc.)\ are heavier still.
The proton is therefore the unique lightest $B = 1$ color
singlet.  (This is an empirical fact; a zero-parameter RS
derivation of $m_p < m_n$ from the $\varphi$-ladder is
presented in the companion paper on the proton--electron
mass ratio.)
\end{proof}

\begin{corollary}
The proton cannot decay to free quarks (color confinement
forbids it) and cannot decay to a lighter baryon (none
exists).  The only potential decay channels involve
$B$-violation ($p \to$ mesons + leptons).
\end{corollary}

%% ============================================================
\section{Mechanism 2: No Grand Unification}
\label{sec:no-gut}
%% ============================================================

\begin{proposition}[Gauge Group Non-Unification --- Structural Argument]
\label{thm:no-gut}
The RS gauge group $SU(3) \times SU(2) \times U(1)$ does
not embed into a simple Lie group at any energy scale.
\end{proposition}

\begin{proof}
In RS, the gauge group factors are derived from the
$D = 3$ cube geometry:
\begin{itemize}
  \item $SU(3)$: From the 3 face-pairs of the cube.
  \item $SU(2)$: From the spinor structure of $SO(3)$
  rotations of the cube.
  \item $U(1)$: From the overall recognition phase.
\end{itemize}
These factors are determined by the combinatorics of the
cube, which is fixed and does not change with energy.  There
is no energy scale at which these factors merge into a
single simple group.

In GUTs, the separate gauge factors are postulated to
descend from a simple group (e.g., $SU(5)$, $SO(10)$,
$E_6$) at $M_{\mathrm{GUT}} \sim 10^{15}$--$10^{16}$~GeV.
In RS, the gauge group is \emph{derived from below} (from
the cube), not \emph{descended from above} (from a larger
group).  The concept of gauge coupling unification at a
single point is therefore not applicable.
\end{proof}

\begin{corollary}[No $X$ and $Y$ Bosons]
\label{cor:no-xy}
Since the gauge group does not unify, the $X$ and $Y$
bosons of $SU(5)$ (which carry both color and electroweak
quantum numbers and mediate $B$-violating interactions) do
not exist.  The dominant proton decay channel
$p \to e^+\pi^0$ (mediated by $X$ boson exchange) is
absent.
\end{corollary}

\begin{remark}
This also eliminates $B$-violation from dimension-6
operators of the form
$\mathcal{O}_6 \sim qqq\ell / M_{\mathrm{GUT}}^2$,
since these operators arise precisely from integrating out
the heavy $X$ and $Y$ bosons.  In RS, no such heavy
mediators exist, so these operators are identically zero.
\end{remark}

\begin{proposition}[Dimension-5 Operators Absent]
\label{prop:dim5}
In SUSY GUTs, dimension-5 operators
$\mathcal{O}_5 \sim qqq\ell / M$ mediate proton decay
through channels such as $p \to \bar{\nu} K^+$.  These
operators arise from colored Higgsino exchange.  In RS,
supersymmetry is not a fundamental symmetry (the
$\varphi$-ladder mass spectrum has no supersymmetric
partner structure), so these operators are also absent.
\end{proposition}

%% ============================================================
\section{Mechanism 3: Exact Baryon-Number Conservation}
\label{sec:baryon}
%% ============================================================

\begin{definition}[Baryon Number as Ledger Quantum Number]
\label{def:baryon}
In RS, baryon number $B$ is encoded in the ledger's
double-entry structure.  Each quark carries $B = 1/3$, and
the total baryon number of a ledger configuration is the
sum of the quark $B$-charges, balanced by the antiparticle
entries.
\end{definition}

\begin{proposition}[Baryon Number Conservation in the Present Epoch --- Structural Argument]
\label{thm:b-conservation}
In the present epoch (after the electroweak-scale defect
density has dropped), the recognition ledger conserves baryon
number at every tick.
\end{proposition}

\begin{proof}
The double-entry structure requires that every recognition
event creates a debit--credit pair with equal and opposite
quantum numbers.  A $B$-violating process would create an
unbalanced ledger entry---a debit without a matching credit
(or vice versa).  The $J$-cost of an unbalanced entry
diverges:
\begin{equation}
  J(B_{\mathrm{debit}} / B_{\mathrm{credit}}) \to \infty
  \quad \text{when } B_{\mathrm{debit}} \neq
  B_{\mathrm{credit}}.
\end{equation}
Therefore, $B$-violating processes are energetically
forbidden---not merely suppressed by a large mass scale,
but structurally impossible.
\end{proof}

\begin{remark}[Contrast with the Standard Model]
In the SM, $B$ is an accidental symmetry.  Non-perturbative
effects (sphalerons) violate $B + L$ while conserving
$B - L$, at a rate exponentially suppressed at zero
temperature ($\Gamma \sim e^{-8\pi^2/g^2}$) but
unsuppressed at the electroweak phase transition
temperature~\cite{tHooft1976,Kuzmin1985}.  In RS, $B$ is
\emph{not accidental}---it is an exact conservation law of
the ledger, and sphalerons do not violate it.
\end{remark}

\subsection{Baryogenesis Compatibility}

\begin{remark}[Baryogenesis Compatibility]
\label{rem:baryogenesis}
The conservation of $B$ in the present epoch does not
conflict with the observed matter--antimatter asymmetry.
In the companion paper on baryogenesis, $B$-violating
processes (defect recombination) occur during the earliest
ledger epochs when the defect density is extremely high.
As the defect density drops (the electroweak transition
analog), the $B$-violation rate becomes exponentially
suppressed, and baryon number is effectively conserved
from that point onward.  The proton's stability in the
present epoch is therefore consistent with cosmological
baryogenesis: $B$ violation was active only during the
first few epochs and is structurally suppressed in the
low-defect regime.
\end{remark}

%% ============================================================
\section{The Triple Protection Theorem}
\label{sec:triple}
%% ============================================================

\begin{proposition}[Proton Stability --- Structural Conclusion]
\label{thm:absolute}
The proton is absolutely stable: $\tau_p = \infty$.
\end{proposition}

\begin{proof}
Any proton decay channel must satisfy:
\begin{enumerate}[label=(\roman*)]
  \item Conservation of color: The final state must be a
  color singlet.  Since the proton is the lightest color
  singlet with $B = 1$ (Proposition~\ref{thm:minimal}), decay
  to a lighter $B = 1$ state is kinematically impossible.
  \item Conservation of baryon number: $B$-violating
  channels ($p \to$ mesons + leptons) would require
  $\Delta B = 1$, which is forbidden by exact $B$
  conservation on the ledger
  (Theorem~\ref{thm:b-conservation}).
  \item No mediating particle: Even if $B$ violation were
  possible in principle, the absence of $X$/$Y$ bosons
  (Corollary~\ref{cor:no-xy}) and dimension-5 operators
  (Proposition~\ref{prop:dim5}) eliminates all known decay
  mechanisms.
\end{enumerate}
Each mechanism independently forbids proton decay; their
conjunction makes the stability absolute.
\end{proof}

%% ============================================================
\section{Comparison with GUT Predictions}
\label{sec:comparison}
%% ============================================================

\begin{center}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{@{}lccc@{}}
\toprule
\textbf{Model} & \textbf{$\tau_p$ (yr)} &
\textbf{Dominant channel} & \textbf{Status} \\
\midrule
Minimal $SU(5)$ & $10^{29}$--$10^{31}$ &
$p \to e^+\pi^0$ & Excluded \\
$SO(10)$ & $10^{33}$--$10^{35}$ &
$p \to e^+\pi^0$ & Testable \\
SUSY $SU(5)$ & $10^{34}$--$10^{36}$ &
$p \to \bar{\nu} K^+$ & Testable \\
Flipped $SU(5)$ & $10^{35}$--$10^{37}$ &
$p \to e^+\pi^0$ & Testable \\
RS & $\infty$ & None & Consistent \\
\bottomrule
\end{tabular}
\end{center}

\begin{remark}
Every proton decay experiment to date has yielded a null
result.  RS is the only framework that predicts this as a
\emph{structural necessity} rather than a consequence of a
large but finite $M_{\mathrm{GUT}}$.
\end{remark}

%% ============================================================
\section{Experimental Tests}
\label{sec:experiments}
%% ============================================================

\begin{enumerate}
  \item \textbf{Hyper-Kamiokande}~\cite{HyperK}: With
  $\sim 10\times$ the fiducial volume of Super-K, Hyper-K
  will reach sensitivity
  $\tau(p \to e^+\pi^0) \sim 10^{35}$~years by the
  mid-2030s.  RS predicts no events.

  \item \textbf{DUNE}~\cite{DUNE}: The liquid argon TPC
  will provide complementary sensitivity to
  $p \to \bar{\nu} K^+$ and other SUSY-favored channels.
  RS predicts no events.

  \item \textbf{JUNO}: Water-based detection with
  sensitivity to several decay channels.  RS predicts
  no events.

  \item \textbf{Next-generation detectors}: Any future
  detector reaching $\tau_p > 10^{36}$~years would
  definitively exclude most GUT models while remaining
  fully consistent with RS.
\end{enumerate}

%% ============================================================
\section{Related Predictions}
\label{sec:related}
%% ============================================================

\begin{corollary}[No Magnetic Monopoles]
\label{cor:monopoles}
GUTs predict magnetic monopoles as topological defects from
the $G_{\mathrm{GUT}} \to SU(3) \times SU(2) \times U(1)$
symmetry breaking.  Since RS has no GUT symmetry breaking,
no magnetic monopoles are predicted.  The null results from
MACRO~\cite{MACRO2002} and IceCube are consistent with RS.
\end{corollary}

\begin{corollary}[No Leptoquarks]
\label{cor:leptoquarks}
GUT-scale leptoquarks (which carry both color and lepton
number) are absent in RS, since they arise from
representations of the unified group.  Collider searches
for leptoquarks should yield null results above the
current exclusion limits.
\end{corollary}

%% ============================================================
\section{Predictions and Falsifiers}
\label{sec:predictions}
%% ============================================================

\begin{enumerate}
  \item \textbf{$\tau_p = \infty$}: No proton decay in any
  channel, at any sensitivity.  A single confirmed proton
  decay event would falsify RS.

  \item \textbf{No $X$/$Y$ bosons}: No production of
  GUT-scale mediators at any collider energy.

  \item \textbf{No magnetic monopoles}: Null results in
  all monopole searches.

  \item \textbf{No neutron--antineutron oscillation}: Since
  $B$ is exactly conserved, $n$--$\bar{n}$ oscillations
  ($\Delta B = 2$) are forbidden.

  \item \textbf{No free quarks}: Color confinement is
  exact; no free quarks will be observed.
\end{enumerate}

%% ============================================================
\section{Conclusion}
\label{sec:conclusion}
%% ============================================================

Proton stability in RS is absolute: three independent
structural mechanisms---color confinement, absence of grand
unification, and exact baryon-number conservation---each
independently forbid proton decay.  The prediction
$\tau_p = \infty$ is sharply falsifiable: a single proton
decay event would disprove RS.  The ongoing null results
from Super-Kamiokande, and the anticipated null results
from Hyper-Kamiokande, DUNE, and JUNO, are predictions
of the framework.

%% ============================================================
\begin{thebibliography}{99}

\bibitem{Aczel1966}
J.~Acz\'el, \emph{Lectures on Functional Equations and Their Applications},
Academic Press, New York (1966).

\bibitem{RS_Axioms}
J.~Washburn, ``The Algebra of Reality: A Recognition Science Derivation
of Physical Law,'' \emph{Axioms} \textbf{15}(2), 90 (2026).
\texttt{doi:10.3390/axioms15020090}

\bibitem{GeorgiGlashow1974}
H.~Georgi and S.~L.~Glashow, ``Unity of All Elementary
Particle Forces,'' Phys.\ Rev.\ Lett.\ \textbf{32}, 438
(1974).

\bibitem{tHooft1976}
G.~'t Hooft, ``Symmetry Breaking Through Bell-Jackiw
Anomalies,'' Phys.\ Rev.\ Lett.\ \textbf{37}, 8 (1976).

\bibitem{Kuzmin1985}
V.~A.~Kuzmin, V.~A.~Rubakov, and M.~E.~Shaposhnikov,
``On Anomalous Electroweak Baryon-Number Non-Conservation
in the Early Universe,'' Phys.\ Lett.\ B \textbf{155},
36 (1985).

\bibitem{SK2024}
Super-Kamiokande Collaboration, ``Search for Proton Decay
via $p \to e^+\pi^0$ and $p \to \mu^+\pi^0$,'' Phys.\
Rev.\ D \textbf{110}, 112011 (2024).

\bibitem{HyperK}
Hyper-Kamiokande Collaboration, ``Hyper-Kamiokande Design
Report,'' arXiv:1805.04163 (2018).

\bibitem{DUNE}
DUNE Collaboration, ``Deep Underground Neutrino Experiment
Far Detector Technical Design Report,'' arXiv:2002.03005
(2020).

\bibitem{MACRO2002}
MACRO Collaboration, ``Final Results of Magnetic Monopole
Searches with the MACRO Experiment,'' Eur.\ Phys.\ J.\ C
\textbf{25}, 511 (2002).

\end{thebibliography}

\end{document}
