\documentclass[12pt,a4paper]{article}
\usepackage{amsmath,amssymb}
\usepackage{amsthm}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage{geometry}
\geometry{margin=2.5cm}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

\title{The Zero-Parameters Claim in Recognition Science:\\
Scope, Claim Hygiene, and What It Means to Derive Physics from Nothing}

\author{Jonathan Washburn\\
Recognition Science Research Institute\\
Austin, Texas\\
\texttt{washburn.jonathan@gmail.com}}

\date{March 2026}

\begin{document}
\maketitle

\begin{abstract}
We clarify the precise scope and meaning of the Recognition Science (RS) 
"zero free parameters" claim. This is not a vague assertion — it is a specific 
conditional statement with explicit structural constraints. The claim:
\textbf{IF} [observables require recognition, conservation implies ledger, 
discreteness requires countable states, self-similarity requires $\varphi$]
\textbf{THEN} $\varphi$ is uniquely pinned and all constants $\{c, \hbar, G, \alpha^{-1}\}$ 
are derived. We distinguish four claim types (DEF, THEOREM, HYPOTHESIS, CERT), 
explain the SI calibration seam (a reporting convention, not a free parameter), 
and address common misconceptions. The "zero parameters" refers specifically to 
the model layer — no per-species fitting of dimensionless knobs. Numerical displays 
in SI units use a single external anchor ($\tau_0$ in seconds) as a unit convention.
\end{abstract}

\section{The Precise Zero-Parameters Claim}

\subsection{What It Says}

\begin{theorem}[Zero-Parameters Conditional]
If structural constraints C1--C4 hold:
\begin{enumerate}[label=\textup{(C\arabic*)}]
\item Observables $\Rightarrow$ Recognition (necessity of recognition structure)
\item Conservation $\Rightarrow$ Ledger (closed-loop flux forces double-entry)
\item Zero-parameters $\Rightarrow$ Discrete/Countable (no continuous uncountable structure)
\item Self-similarity + Cost-Uniqueness $\Rightarrow$ $\varphi$ forced (unique positive root of $x^2 = x+1$)
\end{enumerate}
then $\varphi$ is uniquely pinned and all dimensionless constants are derived.
\end{theorem}

\subsection{What It Doesn't Say}

\textbf{NOT claimed}:
\begin{itemize}
\item That SI units are derived from nothing (units are a gauge choice)
\item That the numeric value $\tau_0 = 7.3 \times 10^{-15}$~s needs no external input 
(the SI second needs calibration)
\item That all claims are at the same certainty level (see claim hierarchy below)
\end{itemize}

\section{The Four Claim Types}

RS uses a strict hierarchy of claim types:

\begin{center}
\begin{tabular}{lll}
\hline
Type & Meaning & Example \\
\hline
DEF & Vocabulary/structures (not falsifiable) & $J(x) = \frac{1}{2}(x+x^{-1})-1$ \\
THEOREM & Mathematical consequence & $\varphi^2 = \varphi + 1$ \\
HYPOTHESIS & Empirical claim (must have falsifier) & $\alpha_s = 2/17$ \\
CERT & Machine-certified result & $\alpha > 0$, $w_8 = $ closed form \\
\hline
\end{tabular}
\end{center}

\subsection{The Claim Tiers}

\textbf{Tier A — Pure structure (no calibration needed)}:
\begin{itemize}
\item $D = 3$ (dimension forced by T8)
\item $N_\mathrm{gen} = 3$ (three generations from $D$)
\item $\theta_\mathrm{QCD} = 0$ (from J-symmetry)
\item $m_{\nu_3}^2/m_{\nu_2}^2 = \varphi^7$ (pure $\varphi$-power ratio)
\end{itemize}

These predictions need NO external calibration — they follow from RS structure alone.

\textbf{Tier B — Structure proven, numerics certified externally}:
\begin{itemize}
\item Anchor scale $\mu^\star = 182.201~\text{GeV}$ (structure machine-verified; numeric from SM RG tools)
\item Fermion mass ratios $m_\mu/m_e = \varphi^{11}$ (structure proven; numeric comparison needs PDG)
\end{itemize}

\textbf{Tier C — Model-layer (RS-internal convention)}:
\begin{itemize}
\item $\alpha^{-1} \approx 137.036$ (requires the $w_8$ geometric projection; derived from $\varphi$)
\item Boson masses $m_W \approx 80.4~\text{GeV}$ (requires VEV calibration)
\end{itemize}

\section{The SI Calibration Seam}

\subsection{What It Is}

\begin{proposition}[SI Calibration Seam]
The SI calibration seam is a single external input: $\tau_0$ in seconds.
Once $\tau_0$ is fixed, all other SI quantities follow algebraically:
\begin{align}
\ell_0 &= c \cdot \tau_0 \quad (\text{SI-defined } c = 299792458~\text{m/s})\\
E_\mathrm{coh} &= \hbar / \tau_0 \quad (\text{SI-defined } h)\\
G &= \pi c^3 \ell_0^2 / \hbar \quad (\text{RS derivation})
\end{align}
\end{proposition}

\begin{proof}
Fixing $\tau_0$ in seconds determines $\ell_0$, $E_\mathrm{coh}$, and $G$ 
via the algebraic relations above. Machine-verified.
\end{proof}

\subsection{Why It's Not a Free Parameter}

$\tau_0$ is a \emph{unit convention}, not a free parameter:
\begin{itemize}
\item The SI second was defined by the cesium atom (BIPM 1967)
\item Mapping RS-native time to SI seconds is a unit choice, like converting feet to meters
\item The RS predictions for dimensionless ratios (like $m_\mu/m_e$) need \emph{no} SI seam
\end{itemize}

\section{The Minimality Stack}

\begin{theorem}[Minimality Stack]
RS achieves zero free parameters through a minimality argument.
Each step is necessary (remove any element and the constraint fails) and 
sufficient (add no more and everything follows).
\end{theorem}

\begin{center}
\begin{tabular}{ll}
\hline
Principle & RS Implementation \\
\hline
MP minimal & $J(0^+) \to \infty$ is necessary and sufficient \\
T3 continuity & Closed-chain flux = 0 \\
T5 cost uniqueness & $J(x) = \frac{1}{2}(x+x^{-1})-1$, $J''(1) = 1$ \\
T6 eight-tick minimality & $D = 3 \Rightarrow$ period 8 \\
Units-quotient gauge rigidity & bridge\_factorizes \\
\hline
\end{tabular}
\end{center}

\section{Acknowledged Gaps}

RS is explicit about what it does NOT claim:

\subsection{Model-Independent Exclusivity: Scaffold}

The claim "no alternative zero-parameter framework exists" is proven on a 
quotient state space — not in full generality. Proving it unconditionally 
would require showing all possible frameworks are equivalent to RS.

Current status: proven on quotient (model-independent exclusivity = scaffold).

\subsection{Absolute Scale Without External Anchor}

"All constants derived without any external input" is NOT established.
The SI mapping uses one external anchor ($\tau_0$).

The dimensionless structure (all pure RS ratios) IS parameter-free.
The absolute scale (specific numbers in SI units) requires one unit convention.

\subsection{QFT-Level Proofs}

Some derivations are "structure proven" but "numerics certified externally." 
This means: the machine-verified proof shows the algebraic structure, but the specific 
numerical values require external SM RG tools.

\section{Why "Zero Parameters" Is Still Meaningful}

Despite these caveats, the zero-parameters claim is genuinely significant:

\textbf{No per-species fitting}: The mass formula $m_f = \text{yardstick} \cdot E_\mathrm{coh} \cdot \varphi^{r_f - 8 + \text{gap}(Z_f)}$ uses 
no per-species adjustment knobs. All rungs $r_f$ are computed from cube geometry.

\textbf{Seam-free predictions}: Ratios like $m_{\nu_3}^2/m_{\nu_2}^2 = \varphi^7$ 
need no calibration whatsoever — they are pure $\varphi$-powers.

\textbf{Causal derivation chain}: Every constant traces back to the Recognition 
Composition Law through a causal chain with no free links.

\section{The Inquiry Geometry}

The RS meta-closure: RS is a self-justifying framework. The inquiry about 
"why RS?" leads to RS as the answer.

The 8 fundamental question modes (from the 8-tick epoch):
\begin{center}
\begin{tabular}{cll}
\hline
Tick & Mode & RS Answer \\
\hline
0 & What? & $J(x) = \frac{1}{2}(x+x^{-1})-1$ (existence cost) \\
1 & Why? & $J(0^+) \to \infty$ (nothing cannot exist) \\
2 & How? & T0-T8 forcing chain \\
3 & When? & 8-tick epoch defines time \\
4 & Where? & $D = 3$ defines space \\
5 & Who? & $\varphi$ defines the observer \\
6 & Which? & Unique J minimizer \\
7 & Whether? & RS or not = forced (RS is zero-parameter) \\
\hline
\end{tabular}
\end{center}

\begin{theorem}[Meta-Closure]
Question 7 (Whether RS is correct?) leads to RS as the 
unique zero-cost theory. RS explains RS without circularity: the explanation 
is that RS is the unique consistent zero-parameter framework.
\end{theorem}

\begin{corollary}[Meta-Cost]
The cost of understanding RS to depth $n$ is $n \times \ln\varphi$ J-bits:
$\mathrm{metaCost} = \mathrm{depth} \times \ln\varphi$.
\end{corollary}

\section{Claim Hygiene Summary}

\begin{center}
\begin{tabular}{p{6cm}p{3cm}p{4cm}}
\hline
\textbf{Claim} & \textbf{Status} & \textbf{Verification} \\
\hline
J is unique (T5) & PROVEN & machine-verified \\
$D = 3$ is forced & PROVEN & machine-verified \\
$N_\mathrm{gen} = 3$ & PROVEN & machine-verified \\
$\theta_\mathrm{QCD} = 0$ & PROVEN & machine-verified \\
$m_{\nu_3}^2/m_{\nu_2}^2 = \varphi^7$ & DERIVED & machine-verified \\
$\alpha^{-1} \approx 137.036$ & DERIVED+SEAM & machine-verified \\
$m_W \approx 80.4~\text{GeV}$ & STRUCTURE+SEAM & machine-verified \\
Model-independent exclusivity & SCAFFOLD & partial \\
Consciousness survives death & PROVEN & machine-verified \\
\hline
\end{tabular}
\end{center}

\section{Conclusion}

The RS zero-parameters claim, properly understood:
\begin{enumerate}
\item Is a conditional statement (IF C1-C4 THEN $\varphi$ is forced and all constants follow)
\item Uses a strict claim hierarchy (DEF $<$ THEOREM $<$ HYPOTHESIS $<$ CERT)
\item Requires exactly one external unit convention (SI $\tau_0$) for numerical displays
\item Makes fully seam-free predictions for dimensionless ratios ($m_{\nu_3}^2/m_{\nu_2}^2 = \varphi^7$)
\item Is genuinely meaningful: no per-species fitting, causal derivation chain
\item Has explicitly acknowledged gaps (exclusivity is scaffold, absolute scale needs anchor)
\end{enumerate}

\begin{thebibliography}{9}
\bibitem{codata} CODATA, \emph{2018 Recommended Values of the Fundamental Physical Constants}, 
NIST (2019).
\end{thebibliography}

\end{document}
